---
title: "Lesson 1: Types of Data, Sampling & Study Design"
subtitle: "MA206X: Probability and Statistics"
date: today
format:
  html:
    code-fold: show
---

```{r setup, include=FALSE}
library(tidyverse)
library(kableExtra)
```

::: {.learning-objectives}
### Learning Objectives

By the end of this lesson, you will be able to:

1. Distinguish populations, samples, and processes in studies
2. Classify variables as categorical or quantitative
3. Identify parameters vs. statistics in context

**Reading:** Devore 1.1
:::

## Introduction

Statistical thinking begins with understanding **what** we're measuring and **how** we collect that information. Before we can analyze data, we must clearly define:

- What entities we're studying (observational units)
- What characteristics we're measuring (variables)
- Whether we're studying everyone or just a subset (population vs. sample)

## Populations, Samples, and Processes

### Key Definitions

::: {.callout-note}
## Fundamental Concepts

**Population**: The entire collection of individuals or objects about which information is desired.

**Sample**: A subset of the population selected for study.

**Parameter**: A numerical characteristic of a **population** (usually unknown).

**Statistic**: A numerical characteristic of a **sample** (computed from data).
:::

### Why Study Samples?

In most real-world scenarios, studying an entire population is:

- **Impractical** - Too time-consuming or expensive
- **Impossible** - Population is infinite or inaccessible
- **Unnecessary** - A well-chosen sample provides reliable information

The key challenge: **How do we ensure our sample accurately represents the population?**

::: {.example}
### Example 1.1: Population vs. Sample

**Scenario**: You want to know the average height of all cadets in the Corps.

- **Population**: All 4,400 cadets at USMA
- **Sample**: 200 randomly selected cadets
- **Parameter**: $\mu$ = true average height of all cadets (unknown)
- **Statistic**: $\bar{x}$ = average height of the 200 sampled cadets (computable)

If we calculate $\bar{x} = 69.5$ inches from our sample, we **estimate** that $\mu \approx 69.5$ inches.
:::

## Types of Variables

Variables are characteristics that can differ from one observational unit to another. Understanding variable types determines which statistical methods we can use.

### Categorical Variables

**Categorical** (qualitative) variables place individuals into groups or categories.

#### Subtypes:

1. **Nominal**: Categories with no natural ordering
   - Examples: Color (red, blue, green), Military Occupational Specialty (11A, 35D, 12A)

2. **Ordinal**: Categories with a meaningful order
   - Examples: Class rank (4th, 3rd, 2nd, 1st class), Likert scale (Strongly Disagree → Strongly Agree)

3. **Binary**: Only two possible categories
   - Examples: Pass/Fail, Yes/No, True/False

### Quantitative Variables

**Quantitative** (numerical) variables take numerical values where arithmetic operations make sense.

#### Subtypes:

1. **Discrete**: Countable values (often integers)
   - Examples: Number of pushups, number of siblings, count of vehicles

2. **Continuous**: Can take any value in an interval
   - Examples: Height, weight, temperature, time

### Quick Classification Guide

```{r, echo=FALSE}
var_guide <- tibble(
  Question = c(
    "Can you perform arithmetic?",
    "Are values countable?",
    "Is there a natural order?"
  ),
  `If Yes` = c(
    "Quantitative",
    "Discrete",
    "Ordinal"
  ),
  `If No` = c(
    "Categorical",
    "Continuous",
    "Nominal"
  )
)

var_guide %>%
  kable(align = "c") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

::: {.example}
### Example 1.2: Classifying Variables

Let's classify variables from a cadet fitness study:

```{r, echo=FALSE}
fitness_vars <- tibble(
  Variable = c(
    "APFT Score",
    "Gender",
    "Company",
    "2-Mile Run Time (minutes)",
    "Number of Pull-ups",
    "Fitness Category (Excellent/Good/Fair/Poor)"
  ),
  Type = c(
    "Quantitative (Discrete)",
    "Categorical (Binary)",
    "Categorical (Nominal)",
    "Quantitative (Continuous)",
    "Quantitative (Discrete)",
    "Categorical (Ordinal)"
  ),
  Reasoning = c(
    "Counts total points (0-300)",
    "Two categories: Male/Female",
    "No natural ordering of A-I companies",
    "Can take any positive value",
    "Count of repetitions",
    "Ordered quality levels"
  )
)

fitness_vars %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  column_spec(1, bold = TRUE)
```
:::

## Working with Real Data in R

Let's explore variable types using a dataset. We'll use R's built-in `mtcars` dataset about automobiles.

```{r}
# Load the dataset
data(mtcars)

# View the first few rows
head(mtcars)
```

### Identifying Variable Types

```{r}
# Structure of the dataset
str(mtcars)
```

Let's classify each variable:

```{r}
# Create a summary of variable types
mtcars_summary <- tibble(
  Variable = names(mtcars),
  Type = c(
    "Quantitative (Continuous)",  # mpg
    "Quantitative (Discrete)",     # cyl
    "Quantitative (Continuous)",   # disp
    "Quantitative (Continuous)",   # hp
    "Quantitative (Continuous)",   # drat
    "Quantitative (Continuous)",   # wt
    "Quantitative (Continuous)",   # qsec
    "Categorical (Binary)",        # vs (engine type)
    "Categorical (Binary)",        # am (transmission)
    "Quantitative (Discrete)",     # gear
    "Quantitative (Discrete)"      # carb
  )
)

mtcars_summary %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

### Converting Variables in R

Sometimes R doesn't automatically recognize categorical variables. We can convert them:

```{r}
# Convert categorical variables to factors
mtcars_clean <- mtcars %>%
  mutate(
    cyl = factor(cyl),
    vs = factor(vs, labels = c("V-shaped", "Straight")),
    am = factor(am, labels = c("Automatic", "Manual")),
    gear = factor(gear),
    carb = factor(carb)
  )

# Check the structure again
str(mtcars_clean)
```

## Parameters vs. Statistics: Notation Matters

Statistical inference involves using sample statistics to make claims about population parameters. We use different notation for each:

```{r, echo=FALSE}
notation_table <- tibble(
  Measure = c(
    "Mean",
    "Proportion",
    "Standard Deviation",
    "Variance",
    "Correlation",
    "Size"
  ),
  Population = c(
    "$\\mu$ (mu)",
    "$\\pi$ or $p$",
    "$\\sigma$ (sigma)",
    "$\\sigma^2$",
    "$\\rho$ (rho)",
    "$N$"
  ),
  Sample = c(
    "$\\bar{x}$ (x-bar)",
    "$\\hat{p}$ (p-hat)",
    "$s$",
    "$s^2$",
    "$r$",
    "$n$"
  )
)

notation_table %>%
  kable(escape = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover")) %>%
  column_spec(2, background = "#e3f2fd") %>%
  column_spec(3, background = "#fff3e0")
```

::: {.callout-warning}
## Common Mistake

Don't confuse **population proportion** $p$ with **p-value**! Context matters.

- When discussing "true proportion of successes" → population parameter $p$ or $\pi$
- When discussing "probability of observing data if null hypothesis is true" → p-value
:::

::: {.example}
### Example 1.3: Parameters and Statistics

**Scenario**: The Army wants to know what percentage of all soldiers prefer virtual training to in-person training.

- **Population**: All active-duty soldiers
- **Parameter of interest**: $\pi$ = proportion of ALL soldiers who prefer virtual training
- **Sample**: 1,000 randomly selected soldiers
- **Statistic**: $\hat{p} = 0.62$ (62% of sampled soldiers prefer virtual)

**Inference**: We use $\hat{p} = 0.62$ to estimate that $\pi \approx 0.62$.

Later in the course, we'll learn how to quantify our uncertainty about this estimate using **confidence intervals**.
:::

## Practice Problems

::: {.practice}
### Practice 1.1

For each scenario, identify:

a) The population
b) The sample (if applicable)
c) The parameter of interest
d) Any statistics mentioned

**Scenario 1**: A study of 500 West Point graduates found that 73% were promoted to Major within 10 years of commissioning.

**Scenario 2**: The registrar wants to know the average GPA of all cadets in the Class of 2026.

**Scenario 3**: All members of a particular company took the ACFT. The company commander calculates the average score.
:::

::: {.practice}
### Practice 1.2

Classify each variable as categorical or quantitative. If categorical, specify nominal/ordinal/binary. If quantitative, specify discrete/continuous.

a) Rank (CPT, MAJ, LTC, COL)
b) Blood pressure (mmHg)
c) Number of deployments
d) Branch (Infantry, Armor, Signal, etc.)
e) Ruck march time (minutes)
f) Graduating class rank (1-1000)
g) State of legal residence
h) Combat patch (Yes/No)
:::

## Summary

In this lesson, we established the foundation for statistical thinking:

- **Populations** contain all individuals of interest; **samples** are subsets we actually study
- **Parameters** describe populations (unknown); **statistics** describe samples (computed from data)
- **Categorical variables** group individuals into categories (nominal, ordinal, binary)
- **Quantitative variables** measure numerical quantities (discrete, continuous)

Understanding these distinctions is critical for:

- Choosing appropriate statistical methods
- Correctly interpreting results
- Communicating findings clearly

In Lesson 2, we'll explore **how** to collect samples and design studies to ensure our conclusions are valid.

## Additional Resources

- Devore Section 1.1
- [R for Data Science - Data Transformation](https://r4ds.had.co.nz/transform.html)
- Practice problems in WebAssign
